Многоспиновая запутанность в многоквантовом ЯМР с диполярно упорядоченным начальным состоянием

### 2020-05-27 maniscript_rus.tex
Add manuscript translation to Russian.

\documentclass[review]{elsarticle}

@@

\title{Many-spin entanglement in multiple quantum NMR with a dipolar ordered initial state}

\author[icp]{E.~B.~Feldman}
\author[icp,msu]{I.~D.~Lazarev} %

\address[icp]{Institute of Problems of Chemical Physics of Russian Academy of Sciences, \\ Chernogolovka, Moscow Region, Russia 142432}
\address[msu]{Faculty of Fundamental Physical-Chemical Engineering, Lomonosov Moscow State University, GSP-1, Moscow, Russia 119991}

\begin{abstract}
Multiple quantum (MQ) NMR with a dipolar ordered initial state opens new possibilities for the exploration of many-spin entanglement. 
In this paper, we investigate many-spin entanglement in a gas of spin-carrying molecules (atoms) in nanocavities 
in the conditions of MQ NMR with a dipolar ordered initial state.
The second moment of the distribution of the intensities of MQ NMR coherences,
which provides a lower bound on the quantum Fisher information, 
is used for an estimate of the number of the entangled spins. 
Many-spin entanglement is investigated at different temperatures and different numbers of spins.
\end{abstract}

\begin{keyword}
multiple quantum (MQ) NMR \sep  
quantum correlations \sep 
quantum Fisher information \sep 
entanglement, nano-pore \sep 
MQ NMR coherence \sep 
second moment, temperature \sep 
dipolar ordered state \sep 
two-pulse Broekaert-Jeener sequence
\end{keyword}

@@

\linenumbers

@@

Entanglement~\cite{Nielsen_2009} is an important notion in quantum mechanics. It is responsible, in particular, for advantages of quantum computers over their classical counterparts.
Recently demonstrated quantum supremacy of a programmable superconducting processor~\cite{Arute2019} is also connected with entanglement, which is absent in classical physics.
Among numerous methods of the investigation of entanglement, we focus on multiple quantum (MQ) NMR in solids~\cite{Baum_1985}, which is widely used to characterize entanglement in binary systems~\cite{Furman_2008,Furman_2009,Fel_dman_2008,Fel_dman_2012}. 
It turns out that the MQ NMR spectroscopy~\cite{Baum_1985} allows us to extract information about many-spin entanglement~\cite{G_rttner_2018} by using the quantum Fisher information~\cite{T_th_2014,Pezz__2018}.

The quantum Fisher information describes the quickness of a change of quantum states determined by a density matrix in response to a change of some parameter. 
In MQ NMR spectroscopy, that parameter is the phase increment between the radio-frequency (rf) pulses, irradiating the system on the preparation and mixing periods of the MQ NMR experiment~\cite{Baum_1985} 
that leads to the separation of the signals, corresponding to MQ NMR coherences of different orders, and allows obtaining the total MQ NMR spectrum. 
The phase increment is proportional to the duration of the evolution period.
Using the quantum Fisher information (QFI)~\cite{Liu_2014} for the analysis of the MQ NMR spectra, one can extract important information about many-spin entanglement.
The point is that there is a relation between the second moment of the MQ NMR spectrum~\cite{Khitrin_1997} and the quantum Fisher information~\cite{G_rttner_2018,Doronin_2019}.
Moreover, the second moment of the MQ NMR spectrum provides a lower bound on the quantum Fisher information~\cite{G_rttner_2018}.
This means that the MQ NMR spectroscopy is a valuable method for solving quantum information problems.

Many-spin entanglement was investigated~\cite{Doronin_2019} for a nonspherical nanopore filled with a gas of spin-carrying molecules in a strong external magnetic field~\cite{Baugh_2001,Doronin_2009}.
The thermally equilibrium initial state of the system was determined by the one-spin Zeeman interaction with the external magnetic field~\cite{Doronin_2007a}.
It is also possible to investigate many-spin entanglement when the same system is initially prepared in the dipolar ordered state~\cite{Goldman_1970} using either the adiabatic demagnetization method in a rotating reference frame (RRF)~\cite{Goldman_1970,Slichter_1961} or the two-pulse Broekaert-Jeener sequence~\cite{Goldman_1970,Jeener_1967}.
The MQ NMR dynamics with this initial state have been simulated both in small spin systems~\cite{Doronin_2007a,Doronin_2007b} and in a system consisting of 200-600 spin-carrying molecules (atoms) filling a nanopore~\cite{Doronin_2011}.
The approaches developed for those investigations are restricted to  high temperatures and cannot be applied to  many-spin entanglement.

In the present article we consider the intermediate-temperature case with low Zeeman temperatures and high dipolar ones. 
Nuclear magnetic ordering~\cite{Abragam_1982} is beyond the scope of this paper.
Notice that the two-pulse Broekaert-Jeneer experiment~\cite{Jeener_1967} was performed in the high-temperature case.
We prove theoretically that the experiment~\cite{Jeener_1967} can be realized also for the intermediate-temperature case. 
It was shown~\cite{Doronin_2011} that in the MQ NMR experiment with the dipolar ordered initial state, MQ NMR coherences emerge faster 
than in the MQ NMR experiment with the thermal-equilibrium initial state in a strong external magnetic field.
This observation is important for many-spin entanglement investigations, because they involve calculations of the second moment of the distribution of MQ NMR coherences. 
It is also important for the investigation of correlation spreading~\cite{Baugh_2001,Baum_1986,S_nchez_2014,Munowitz_1987} and localization~\cite{Alvarez_2015,Wei_2018}.
Indeed, the spreading rate can be described through out-of-time ordered correlations, which are connected with the distribution of MQ NMR coherences.

The present paper investigates many-spin entanglement using the MQ NMR spectrum of spin-carrying atoms (molecules) in a nanopore when the system is prepared in a dipolar ordered state.
In Sec.~\ref{sec:2}, the theory of MQ NMR dynamics at a low Zeeman temperature and a high dipolar temperature is developed.
An analytical solution for the MQ NMR dynamics of a three-spin system is obtained at such temperatures in Sec.~\ref{sec:3}.
The second moment of the MQ NMR spectrum as a measure of many-spin entanglement is considered in Sec.~\ref{sec:4}.
The dependence of many-spin entanglement on the dipolar temperature and the number of the spins in the system is investigated in Sec.~\ref{sec:5}.
We briefly summarize our results in concluding Sec.~\ref{sec:6}.
In the Appendix, we show that the two-pulse Broekaert-Jeener sequence can be used in the case when the Zeeman temperature is low and the dipole one is high.

\section{Theory of MQ NMR dynamics in a nanopore at a low Zeeman temperature and a high dipolar temperature}
\label{sec:2}

MQ NMR dynamics in a nanopore is governed by the Hamiltonian~\cite{Doronin_2019,Doronin_2009} 
%
\begin{equation}
    \label{eq:1}
    H_{\mathrm{MQ}} = - \dfrac{D}{4} \left[
        \left(I^{+}\right)^{2} 
        + \left(I^{-}\right)^{2} 
    \right] ,
\end{equation}
%
where 
%
\begin{equation}
    \label{eq:2}
    I^{\pm} = \sum\limits_{j=1}^{N} I_{j}^{\pm},
\end{equation}
%
$N$ is the number of the spins in the nanopore, $I^{\pm}_{j}$ are the raising or lowering operators of spin $j$, and $D$ is the dipolar coupling constant averaged by the fast molecular diffusion of spin-carrying atoms (molecules) in the nanopore.
We emphasize that the dipolar coupling constant $D$ is the same for all pairs of interacting spins in the nanopore~\cite{Doronin_2019,Doronin_2009}.
The density matrix $\rho(\tau)$ on the preparation period of the MQ NMR experiment~\cite{Baum_1985} can be obtained from the Liouville evolution equation~\cite{Goldman_1970,Abragam_1982} 
%
\begin{equation}
    \label{eq:3}
    i\dfrac{\mathrm{d}\rho(\tau)}{\mathrm{d}\tau} = \left[
    H_\mathrm{MQ},\rho(\tau)
    \right]
\end{equation}
%
with the initial thermodynamic equilibrium density matrix 
%
\begin{equation}
    \label{eq:4}
       \rho(0) = \rho_\mathrm{eq} = \dfrac{1}{Z}
       e^{
		   \frac{\hslash \omega_{0}}{k} \alpha_\mathrm{z} I_\mathrm{z} 
            + \frac{\hslash }{k} \beta_\mathrm{d} H_\mathrm{dz}
        },
\end{equation}
%
where 
$Z = \mathrm{Tr} \left\{ e^{\frac{\hslash \omega_{0}}{k} \alpha_\mathrm{z} I_\mathrm{z} + \frac{\hslash  }{k} \beta_\mathrm{d} H_\mathrm{dz}} \right\}$ is the partition function, 
$\hslash$ and $k$ are the Plank and Boltzmann constants, 
$\omega_{0}$  is the Larmor frequency , $I_\mathrm{z}$ is the operator of the projection of the total spin angular momentum on the z-axis, 
which is directed along the strong external magnetic field,  
$H_\mathrm{dz}$ is the secular part of the dipole-dipole interaction~(DDI) Hamiltonian in a strong external magnetic field, and $\alpha_\mathrm{z}$, $\beta_\mathrm{d}$ are the inverse Zeeman and dipolar temperatures. 
We will consider the case when the Zeeman temperature is low $({\frac{\hslash \omega_{0}}{k} \alpha_\mathrm{z}}\gg 1)$ 
and the dipolar temperature is high $\left( \frac{\hslash{D}}{k}\beta_\mathrm{d} \ll 1\right)$.
For concreteness, we suppose that $\omega_{0} = 2\pi \cdot 500 \cdot 10^{6}$~s$^{-1}$ and $D = 2\pi \cdot 10^{4}$~s$^{-1}$.
In the Appendix, we prove that the two-pulse Broekaert-Jeener sequence~\cite{Goldman_1970,Jeener_1967} results in the dipolar ordered state even at a low Zeeman temperature.
The adiabatic demagnetization~\cite{Goldman_1970,Slichter_1961} is the second method of preparing a system in the dipolar ordered state.
Using those methods, we can obtain the system in the thermodynamic equilibrium state with the density matrix
%
\begin{equation}
    \label{eq:5}
    \rho_i = \frac{1}{Z_i} e^\frac{\hslash\beta_\mathrm{d} \hdz}{k}
    \approx \frac{1}{Z_i}(1 + \frac{\hslash\beta_\mathrm{d}}{k} H_\mathrm{dz}),
\end{equation}
%
where the partition function
%
\begin{equation}
    \label{eq:6}
	Z_i = \mathrm{Tr} \left\{ e^\frac{\hslash\beta_\mathrm{d} \hdz}{k} \right\} \approx 2^{N}.
\end{equation}
%
MQ NMR dynamics in the nanopore will be investigated on the basis of Eq.~(\ref{eq:3}) with the initial state of Eq.~(\ref{eq:5}).
It is also significant that the Hamiltonian $H_{dz}$ is partially averaged by the fast molecular diffusion in the nanopore and the averaged Hamiltonian can be written as \cite{Fel_dman_2004,Doronin_2011}
%
\begin{equation}
    \label{eq:7}
    H_\mathrm{dz} = \dfrac{D}{2} (3 I^{2}_{z} - I^{2}) , % ?
\end{equation}
%
where $I^{2}$ is the square of the spin angular momentum.

Let   $G(\tau,\phi)$  be the signal after the preparation, evolution and mixing periods of the MQ NMR experiment~\cite{Baum_1985}, averaged over the equilibrium density matrix.  $G(\tau,\phi)$ can be written as~\cite{Doronin_2019} 
%
\begin{equation}
    \begin{split}
        \label{eq:8}
        G(\tau,\phi) 
        & = \mathrm{Tr}\left\{
            e^{i H_\mathrm{MQ} \tau} e^{i\phi I_\mathrm{z}} e^{-i H_\mathrm{MQ}\tau} 
            \rho_i 
            e^{i H_\mathrm{MQ} \tau} e^{-i \phi I_\mathrm{z}} e^{-i \phi H_\mathrm{MQ} \tau} 
            \rho_i 
        \right\} \\
        & = \mathrm{Tr} \left\{
        e^{i \phi I_\mathrm{z}}
        \rho(\tau) 
        e^{-i \phi I_\mathrm{z}} 
        \rho(\tau) 
        \right\},
    \end{split}
\end{equation}
%
where
%
\begin{equation}
    \label{eq:9}
    \rho(\tau) 
    = e^{-i H_\mathrm{MQ} \tau } 
    \rho_i 
    e^{i H_\mathrm{MQ} \tau}
\end{equation}
%
is the solution of Eq.~(\ref{eq:3}) at the initial condition of Eq.~(\ref{eq:5}).
It is convenient to expand the spin density matrix, $\rho(\tau)$, in series as
%
\begin{equation}
    \label{eq:10}
    \rho(\tau) = \sum\limits_n \rho_n(\tau),
\end{equation}
%
where $\rho_{n}(\tau)$ is the contribution to $\rho(\tau)$ from the MQ coherence of the $n$-th order~\cite{Fel_dman_1996}.
Then the function $G(\tau,\phi)$ of Eq.~(\ref{eq:8}) can be rewritten as 
%
\begin{equation}
    \label{eq:11}
    G(\tau,\phi) 
    = \sum\limits_n e^{i n \phi} \mathrm{Tr} \left\{ 
        \rho_{n}(\tau) \rho_{-n}(\tau) 
    \right\},
\end{equation}
%
where we took into account that
%
\begin{equation}
    \label{eq:12}
    \left[ I_{\mathrm{z}},\rho_n(\tau) \right] = n \rho_n(\tau)
\end{equation}
%
It is necessary for further calculations to introduce the normalized intensities $J_{n}(\tau)$ $(n=0, \pm 2, \pm 4, \cdots)$ of the MQ NMR coherences
%
\begin{equation}
    \label{eq:13}
    J_{n}(\tau) = \dfrac{\mathrm{Tr} \left\{
    \rho_{n}(\tau) \rho_{-n}(\tau) 
    \right\}} 
    {\mathrm{Tr} \left\{\rho^2_{i} \right\}}
\end{equation}
%
Using Eqs.~(\ref{eq:9}),~(\ref{eq:10}) one can verity that 
%
\begin{multline}
    \label{eq:14}
    \sum\limits_{n} J_{n}(\tau)
    = \dfrac{
        \mathrm{Tr} \left\{
            \sum_{n} \rho_{n}(\tau) \rho_{-n}(\tau)
        \right\}}
    {\mathrm{Tr} \left\{ \rho^2_{i} \right\}} 
    = \dfrac{
        \mathrm{Tr} \left\{
            \sum_{\mathrm{m,n}} \rho_n(\tau)\rho_m(\tau)
    \right\}}
    {\mathrm{Tr} \left\{\rho^2_{i}\right\}}
    \\
    = \dfrac{
        \mathrm{Tr}\left\{\rho^2(\tau)\right\}
    }
    {
        \mathrm{Tr}\left\{\rho^2_{i}(\tau)\right\}
    }
    = \dfrac{
        \mathrm{Tr} \left\{ 
            e^{-i H_\mathrm{MQ} \tau} 
            \rho^{2}_{i}
            e^{i H_\mathrm{MQ} \tau} 
        \right\}
    }
    {
        \mathrm{Tr} \left\{ \rho_{i}^{2} \right\}
    } 
    = 1
\end{multline}
%
One can conclude from Eq.~(\ref{eq:14}) that the sum of the MQ NMR coherences is conserved on the preparation period of the MQ NMR experiment~\cite{Baum_1985}.

The basis consisting of the eigenstates of the operator $I_\mathrm{z}$ (dubbed the multiplicative basis) is widely used for numerical calculations of MQ NMR dynamics~\cite{Zhang_2009}.
Due to the rapid expansion of the Hilbert space with the growth of the number of spins such calculations are possible only for systems with a small number of spins.
That approach is not suitable for investigations of many-spin entanglement.
Since the Hamiltonian $H_{MQ}$ of Eq.~(\ref{eq:1}) commutes with the square of the total spin angular momentum $\hat I^2$, 
it is possible to use the basis consisting of the common eigenstates of $\hat I^2$ and $I_\mathrm{z}$ in order to study MQ NMR dynamics as was done in Ref.~\cite{Doronin_2009,Doronin_2011,Doronin_2019}.
In this basis, the Hamiltonian $H_{MQ}$ and the initial density matrix of Eq.~(\ref{eq:5}) (see also Eq.~(\ref{eq:7})) consist of blocks, corresponding to different values of the spin angular momentum~\cite{Doronin_2009}.
Then the investigation of MQ NMR dynamics can be reduced to solving a set of problems of lower dimensions.

Since the Hamiltonian $H_{MQ}$ of Eq.~(\ref{eq:1}) commutes with the operator $e^{i\pi I_\mathrm{z}}$, the $2^N \times 2^N$ Hamiltonian matrix reduces to two $2^{N-1} \times 2^{N-1}$ submatrices~\cite{Doronin_2009}.
For odd $N$, both submatrices give the same contribution to the MQ NMR coherences, and one should solve the problem using only one $2^{N-1} \times 2^{N-1}$ submatrix and double the obtained intensities. 
In our calculations, we take only odd numbers of spins.
Using this method, one can investigate MQ NMR dynamics in systems consisting of hundreds spins.

\section{Analytical solution for MQ NMR dynamics of a three-spin system in a nanopore in a dipolar ordered state}
\label{sec:3}

\begin{figure}
    \centering
  	\includegraphics[width=0.5\linewidth]{coherences_n3_beta5.eps}
	\caption{
	    Intensities of MQ NMR coherenes $J_{n}$ ($n=0, 2$) in a nanopore with $N=3$.
	}
	\label{fig:1}
\end{figure}

Obtaining the exact solution for MQ NMR dynamics of a three-spin system in a dipolar ordered state in a nanopore is similar to the problem considered in Ref.~\cite{Doronin_2019} for the initial thermodynamic equilibrium in a strong external magnetic field. 
Here we do not use the high temperature approximation~\cite{Goldman_1970}.

>>>>

\documentclass[review]{elsarticle}
\usepackage[english,russian]{babel}

@@

\title{Многоспиновая запутанность в многоквантовом ЯМР с диполярно упорядоченным начальным состоянием}

\author[icp]{Э.~Б.~Фельдман}
\author[icp,msu]{И.~Д.~Лазарев} %

\address[icp]{Институт проблем химической физики РАН, \\ Черноголовка, Московская область, Россия 142432}
\address[msu]{Факультет фундаментальной физико-химической инженерии, МГУ GSP-1, Москва, Россия 119991}

\begin{abstract}
Многоквантовый (МК)  ЯМР с диполярно упорядоченным начальным состоянием открывает новые возможности для изучения многоспиновой запутанности. 
В данной статье исследуется многоспиновая запутанность в газе спин-несущих молекул (атомов) в нанорезонаторах
в условиях ЯМР с дипольно упорядоченным начальным состоянием.
Для оценки количества запутанных спинов используется второй момент распределения интенсивнойстей МК ЯМР-когерентностей, 
который определяет нижнюю границу квантовой информации Фишера. 
Многоспиновая запутанность исследуется при различных температурах и различном количестве спинов.
\end{abstract}

\begin{keyword}
многоквантовый (МК) ЯМР \sep  
квантовые корреляции \sep 
квантовая информация Фишера \sep 
запутанность, нанопора \sep 
МК ЯМР когерентность \sep 
второй момент, температура \sep 
диполярно упорядоченное состояние \sep 
двухимпульсная последовательность Брукерта-Джинера
\end{keyword}

@@

%\linenumbers

@@

Запутанность~\cite{Nielsen_2009} является важным понятием квантовой механики. Наличие данного понятия определеяет преимущество квантовых компьютеров перед их классическими аналогами. 
Продемонстрированное квантовое превосходство программируемого сверхпроводящего процессора~\cite{Arute2019} также связано с понятием запутанности, которое отсутствует в классической физике.
Среди многочисленных методов исследования запутанности мы сосредоточились на многоквантовом (МК) ЯМР в твердых телах~\cite{Baum_1985}, который широко используется для характеристики запутанности в двоичных системах~\cite{Furman_2008, Furman_2009, Fel_dman_2008, Fel_dman_2012}.
МК ЯМР-спектроскопия~\cite{Baum_1985} позволяет извлекать информацию о многоспиновой запутанности~\cite{G_rttner_2018} с помощью квантовой информации Фишера~\cite{T_th_2014, Pezz__2018}.

Квантовая информация Фишера описывает скорость изменения квантовых состояний, определяемых матрицей плотности в ответ на изменение какого-либо параметра.
В МК спектроскопии ЯМР данный параметр представляет собой приращение фазы между радиочастотными импульсами, облучающими систему в периоды приготовления и смешивания в эксперименте по МК ЯМР~\cite{Baum_1985}, 
что приводит к разделению сигналов, соответсвующих МК когерентности ЯМР разных порядков, и позволяет получить полный спектр МК ЯМР. 
Приращение фазы пропорционально продолжительности периода эволюции.
Используя квантовую информацию Фишера~\cite{Liu_2014} для анализа спектра МК ЯМР, можно получить важную информацию о многоспиновой запутанности,
поскольку сущесвует связь между вторым моментом спектра ~\cite{Khitrin_1997} и квантовой информацией Фишера~\cite{G_rttner_2018,Doronin_2019}.
Кроме того второй момент спектра МК ЯМР определяет нижнюю границу квантовой информации Фишера~\cite{G_rttner_2018}.
Это означает, что МК ЯМР-спектроскопия является ценным методом решения квантовых информационных задач.

Многоспиновая запутанность была исследована~\cite{Doronin_2019} для несферической нанопоры, заполненной газом переносящих спин молекул в сильном внешнем магнитном поле~\cite{Baugh_2001,Doronin_2009}.
Термически равновесное начальное состояние системы определялось односпиновым зеемановским взаимодействием с внешним магнитным полем~\cite{Doronin_2007a}.
Также возможно исследовать многоспиновую запутанность, когда та же самая система первоначально подготовлена в диполярном упорядоченном состоянии~\cite{Goldman_1970}, используя либо метод адиабатического размагничивания во вращающейся системе отсчета~\cite{Goldman_1970, Slichter_1961} , либо двухимпульную последовательность Брукерта-Джинера~\cite{Goldman_1970, Jeener_1967}.
Динамика МК ЯМР с этим начальным состоянием моделировалась как в малых спиновых системах~\cite{Doronin_2007a, Doronin_2007b}, так и в системе, состоящей из 200-600 спин-несущих молекул (атомов), заполняющих нанопоры ~\cite{Doronin_2011}.
Однако подходы, разработанные для этих исследований, ограничены высокими температурами и не могут применяться к многоспиновому запутыванию.

В настоящей статье мы рассматриваем среднетемпературный случай с низкими зеемановскими и высокими дипольными.
Магнитное упорядочевание~\cite{Abragam_1982} выходит за рамки данной статьи.
Стоит обратить внимание, что двухимпульсный эксперимент     Брукерта-Дженира~\cite{Jeener_1967} был выполнен при высоких температурах.
В данной статье было доказано, что этот эксперимент~\cite{Jeener_1967} также может выполняться при средних температурах.  
Было показано~\cite{Doronin_2011}, что в МК эксперименте ЯМР с диполярно упорядоченным начальным состоянием МК когерентности ЯМР возникают быстрее, 
чем в МК эксперименте ЯМР с начальным состоянием теплового равновесия в сильном внешнем магнитном поле.
Данное наблюдение является важным в области исследований многоспиновой запутанности, поскольку они используют расчеты второго момента распределения МК когерентностей ЯМР. 
Также оно является значительным в области исследований распространения корреляции~\cite{Baugh_2001,Baum_1986,S_nchez_2014,Munowitz_1987} и локализации~\cite{Alvarez_2015,Wei_2018}.
В действительности, скорость распространения может быть описана с помощью вне временных упорядоченных корреляций, которые связаны с распределением МК когерентностей ЯМР.

В данной статье исследуется многоспиновая запутанность с использованием спектра МК ЯМР спин-несущих атомов (молекул) в нанопоре, когда система подготовлена в диполярно упорядоченном состоянии.
В разделе~\ref{sec:2} описана разработанная теория МК динамики ЯМР при низкой зеемановской температуре и высокой дипольной температуре.
Аналитическое решение для МК динамики ЯМР трехспиновой системы получено при таких же температурах и описано в разеделе~\ref{sec:3}.
Второй момент спектра МК ЯМР в качесвте меры многоспиновой запутанности рассматривается в разделе~\ref{sec:4}.
В разделе~\ref{sec:5} исследуется зависимость многоспиновой запутанности от дипольной температуры и числа спинов в системе.
Краткое изложение всех результатов приводится в разделе~\ref{sec:6}.
В Приложении показано, что двухимпульсная последовательность Брукерта-Джинера может быть использована в случае, когда зеемановская температура низкая, а дипольная - высокая.

\section{Теория МК динамики ЯМР в нанопоре при низкой зеемановской температуре и высокой диполярной}
\label{sec:2}

МК динамика ЯМР в нанопоре определяется гамильтонианом~\cite{Doronin_2019,Doronin_2009} 
%
\begin{equation}
    \label{eq:1}
    H_{\mathrm{MQ}} = - \dfrac{D}{4} \left[
        \left(I^{+}\right)^{2} 
        + \left(I^{-}\right)^{2} 
    \right] ,
\end{equation}
%
где 
%
\begin{equation}
    \label{eq:2}
    I^{\pm} = \sum\limits_{j=1}^{N} I_{j}^{\pm},
\end{equation}
%
$N$ - число спинов в нанопоре, $I^{\pm}_{j}$ - повышающие или понижающие операторы спина $j$, $D$ - константа диполярной связи, усредненная по быстрой молекулярной диффузии спин-несущих атомов (молекул) в нанопоре.
Подчеркнем, что константа диполярной связи $D$ одинакова для всех пар взаимодействующих спинов в нанопоре~\cite{Doronin_2019,Doronin_2009}.
Матрица плотности $\rho(\tau)$ на время подготовки МК эксперимента ЯМР~\cite{Baum_1985} может быть получена из эволюционного уравнения Лиувилля~\cite{Goldman_1970,Abragam_1982} 
%
\begin{equation}
    \label{eq:3}
    i\dfrac{\mathrm{d}\rho(\tau)}{\mathrm{d}\tau} = \left[
    H_\mathrm{MQ},\rho(\tau)
    \right]
\end{equation}
%
с начальной термодинамической матрицей равновесной плотности
%
\begin{equation}
    \label{eq:4}
       \rho(0) = \rho_\mathrm{eq} = \dfrac{1}{Z}
       e^{
            \frac{\hslash \omega}{k} \alpha_\mathrm{z} I_\mathrm{z} 
            + \frac{\hslash }{k} \beta_\mathrm{d} H_\mathrm{dz}
        },
\end{equation}
%
где 
$Z = \mathrm{Tr} \left\{ e^{\frac{\hslash \omega}{k} \alpha_\mathrm{z} I_\mathrm{z} + \frac{\hslash  }{k} \beta_\mathrm{d} H_\mathrm{dz}} \right\}$ - функция разделения, 
$\hslash$ и $k$ - константы Планка и Больцмана, 
$\omega_{0}$ - частота Лармора , $I_\mathrm{z}$ является оператором проекции полного углового момента вращения на ось z, 
который направлен вдоль сильного внешнего магнитного поля,  
$H_\mathrm{dz}$ является секулярной частью гамильтониана диполь-дипольного взаимодействия~(DDI) в сильном внешнем магнитном поле, и $\alpha_\mathrm{z}$, $\beta_\mathrm{d}$ являются обратными зеемановскими и дипольными температурами.
Мы рассмотрим случай, когда температура Зеемана является низкой $({\frac{\hslash \omega}{k} \alpha_\mathrm{z}}\gg 1)$ 
, а диполярная высокой $\left( \frac{\hslash{D}}{k}\beta_\mathrm{d} \ll 1\right)$.
Мы предполагаем, что $\omega_{0} = 2\pi \cdot 500 \cdot 10^{6}$~s$^{-1}$ и $D = 2\pi \cdot 10^{4}$~s$^{-1}$.
В приложении мы доказываем, что двухимпульсная последовательность Брукерта-Джинера~\cite{Goldman_1970,Jeener_1967} приводит к диполярному упорядоченному состоянию даже при низкой зеемановской температуре.
Адиабатическое размагничивание~\cite{Goldman_1970,Slichter_1961} является вторым методом подготовки системы в диполярном упорядоченном состоянии.
Используя эти методы, мы можем получить систему в состоянии термодинамического равновесия с матрицей плотности
%
\begin{equation}
    \label{eq:5}
    \rho_i = \frac{1}{Z_i} e^\frac{\hslash\beta_\mathrm{d} \hdz}{k}
    \approx \frac{1}{Z_i}(1 + \frac{\hslash\beta_\mathrm{d}}{k} H_\mathrm{dz}),
\end{equation}
%
где функция разделения
%
\begin{equation}
    \label{eq:6}
	Z_i = \mathrm{Tr} \left\{ e^\frac{\hslash\beta_\mathrm{d} \hdz}{k} \right\} \approx 2^{N}.
\end{equation}
%
МК динамика ЯМР в нанопоре будет исследована на основе уравнения~(\ref{eq:3}) с начальным состоянием уравнения~(\ref{eq:5}).
Также важно отметить, что гамильтониан $H_{dz}$ частично усредняется быстрой молекулярной диффузией в нанопоре, а усредненный гамильтониан можно записать как \cite{Fel_dman_2004,Doronin_2011}
%
\begin{equation}
    \label{eq:7}
    H_\mathrm{dz} = \dfrac{D}{2} (3 I^{2}_{z} - I^{2}) , % ?
\end{equation}
%
где $I^{2}$ - квадрат углового момента вращения.

Пусть $G(\tau,\phi)$ будет сигналом после периодов подготовки, эволюции и смешивания в МК эксперименте ЯМР~\cite{Baum_1985}, усредненном по матрице равновесной плотности. $G(\tau,\phi)$ может записать как~\cite{Doronin_2019} 
%
\begin{equation}
    \begin{split}
        \label{eq:8}
        G(\tau,\phi) 
        & = \mathrm{Tr}\left\{
            e^{i H_\mathrm{MQ} \tau} e^{i\phi I_\mathrm{z}} e^{-i H_\mathrm{MQ}\tau} 
            \rho_i 
            e^{i H_\mathrm{MQ} \tau} e^{-i \phi I_\mathrm{z}} e^{-i \phi H_\mathrm{MQ} \tau} 
            \rho_i 
        \right\} \\
        & = \mathrm{Tr} \left\{
        e^{i \phi I_\mathrm{z}}
        \rho(\tau) 
        e^{-i \phi I_\mathrm{z}} 
        \rho(\tau) 
        \right\},
    \end{split}
\end{equation}
%
где
%
\begin{equation}
    \label{eq:9}
    \rho(\tau) 
    = e^{-i H_\mathrm{MQ} \tau } 
    \rho_i 
    e^{i H_\mathrm{MQ} \tau}
\end{equation}
%
является решением уравнения~(\ref{eq:3}) при начальном условии уравнения~(\ref{eq:5}).
Матрица спиновой плотности $\rho(\tau)$ раскладывается в ряд как
%
\begin{equation}
    \label{eq:10}
    \rho(\tau) = \sum\limits_n \rho_n(\tau),
\end{equation}
%
где $\rho_{n}(\tau)$ - вклад в $\rho(\tau)$ из МК когерентности $n$-го порядка~\cite{Fel_dman_1996}.
ТОгда функция $G(\tau,\phi)$ уравнения ~(\ref{eq:8}) может быть переписана как
%
\begin{equation}
    \label{eq:11}
    G(\tau,\phi) 
    = \sum\limits_n e^{i n \phi} \mathrm{Tr} \left\{ 
        \rho_{n}(\tau) \rho_{-n}(\tau) 
    \right\},
\end{equation}
%
где мы учитываем, что
%
\begin{equation}
    \label{eq:12}
    \left[ I_{\mathrm{z}},\rho_n(\tau) \right] = n \rho_n(\tau)
\end{equation}
%
Для дальнейших расчетов необходимо ввести нормированные интенсивности $J_{n}(\tau)$ $(n=0, \pm 2, \pm 4, \cdots)$ МК когерентностей ЯМР
%
\begin{equation}
    \label{eq:13}
    J_{n}(\tau) = \dfrac{\mathrm{Tr} \left\{
    \rho_{n}(\tau) \rho_{-n}(\tau) 
    \right\}} 
    {\mathrm{Tr} \left\{\rho^2_{i} \right\}}
\end{equation}
%
Используя уравнения ~(\ref{eq:9}),~(\ref{eq:10}) определим, что
%
\begin{multline}
    \label{eq:14}
    \sum\limits_{n} J_{n}(\tau)
    = \dfrac{
        \mathrm{Tr} \left\{
            \sum_{n} \rho_{n}(\tau) \rho_{-n}(\tau)
        \right\}}
    {\mathrm{Tr} \left\{ \rho^2_{i} \right\}} 
    = \dfrac{
        \mathrm{Tr} \left\{
            \sum_{\mathrm{m,n}} \rho_n(\tau)\rho_m(\tau)
    \right\}}
    {\mathrm{Tr} \left\{\rho^2_{i}\right\}}
    \\
    = \dfrac{
        \mathrm{Tr}\left\{\rho^2(\tau)\right\}
    }
    {
        \mathrm{Tr}\left\{\rho^2_{i}(\tau)\right\}
    }
    = \dfrac{
        \mathrm{Tr} \left\{ 
            e^{-i H_\mathrm{MQ} \tau} 
            \rho^{2}_{i}
            e^{i H_\mathrm{MQ} \tau} 
        \right\}
    }
    {
        \mathrm{Tr} \left\{ \rho_{i}^{2} \right\}
    } 
    = 1
\end{multline}
%
Из уравнения~(\ref{eq:14} можно сделать вывод, что сумма МК когерентнойстей ЯМР сохраняется на период МК эксперимента ЯМР~\cite{Baum_1985}.

Базис, состоящий из собственных состояний оператора $I_\mathrm{z}$ (названный мультипликативным базисом), широко используется для численных расчетов МК динамики ЯМР~\cite{Zhang_2009}.
В связи с быстрым расширением гильбертова пространства с ростом числа спинов такие вычисления возможны только для систем с небольшим числом спинов.
Такой подход не подходит для исследований многоспиновой запутанности.
Поскольку гамильтониан $H_{MQ}$ уравнения ~(\ref{eq:1}) коммутирует с квадратом полного углового момента спина $\hat I^2$, 
можно использовать базис, состоящий из общих собственных состояний $\hat I^2$ и $I_\mathrm{z}$ для изучения МК динамики ЯМР, как это было сделано в статьях ~\cite{Doronin_2009,Doronin_2011,Doronin_2019}.
В этом базисе гамильтониан $H_{MQ}$ и исходная матрица плотности уравнения~(\ref{eq:5}) (см. также уравнение~(\ref{eq:7})) состоят из блоков, соответствующих различным значениям углового момента спина~\cite{Doronin_2009}.
Тогда исследование МК динамики ЯМР может быть сведено к решению ряда задач меньших размеров.

Поскольку гамильтониан $H_{MQ}$ уравнения~(\ref{eq:1}) коммутирует с оператором $e^{i\pi I_\mathrm{z}}$, то $2^N \times 2^N$ гамильтонова матрица сводится к двум подматрицам $2^{N-1} \times 2^{N-1}$ подматрицам~\cite{Doronin_2009}.
Для нечетных $N$ обе подматрицы вносят одинаковый вклад в МК когерентности ЯМР, и нужно решить проблему, используя только одну подматрицу $2^{N-1} \times 2^{N-1}$, и удвоить полученные интенсивности.
В наших расчетах мы берем только нечетные числа спинов.
Используя этот метод, можно исследовать МК динамику ЯМР в системах, состоящих и сотен спинов.

\section{Аналитическое решение для МК динамики ЯМР трехспиновой системы в нанопоре в диполярном упорядоченном состоянии}
\label{sec:3}

\begin{figure}
    \centering
  	\includegraphics[width=0.5\linewidth]{coherences_n3_beta5.eps}
	\caption{
	    Интенсивность МК когерентности ЯМР $J_{n}$ ($n=0, 2$) в нанопоре с $N=3$.
	}
	\label{fig:1}
\end{figure}

Проблема получения точного решения МК динамики ЯМР трехспиновой системы в диполярно упорядоченном состоянии в нанопоре аналогична проблеме, рассмотренной в ~\cite{Doronin_2019} для начального термодинамического равновесия в сильном внешнем магнитном поле.
В работе мы не используем высокотемпературное приближение~\cite{Goldman_1970}.
